---
title: "How to always use kable for data frame output"
date: November 15, 2019
author: Martin Papenberg
---

When writing R Markdown documents and printing a table (e.g., an `R` 
data frame), it is handy to use a formatting function such as 
`knitr::kable()` so that the table looks nicely; otherwise the R 
Markdown output just mimics the `R` console. For example:

```{r}
head(iris)
```

Instead we want:

```{r}
knitr::kable(head(iris))
```

However, I don't want to call `kable()` all the time -- it just clutters 
my code and distracts from the important parts. Fortunately, it is 
possible to define `kable()` (or any other formatting function) as the 
default printing method for data frame (or matrices). This method is 
described in the knitr vignette [Custom Print 
Methods](https://cran.r-project.org/web/packages/knitr/vignettes/knit_print.html). 
To this end, add the following code in an early chunk of your R Markdown 
document:

```{r}
library(knitr) # it is necessary to load knitr
knit_print.data.frame = function(x, ...) {
    res = paste(c("", "", kable(x)), collapse = "\n")
    asis_output(res)
}
registerS3method("knit_print", "data.frame", knit_print.data.frame)
```

Now, if we output a data frame, we get the `kable()` output:

```{r}
head(iris)
```

I am not sure if there is an easier way available, but this works for 
me. 

**EDIT**: There is indeed a much easier way by including the printing 
function in the YAML header (see 
[here](https://bookdown.org/yihui/rmarkdown/pdf-document.html#data-frame 
-printing-1))

```YAML
---
title: "Whatever"
output:
  html_document:
    df_print: kable
---

```

If we also want to print matrices using kable, we could adapt the 
procedure as follows:

```{r}
knit_print.matrix = function(x, ...) {
    res = paste(c("", "", kable(x)), collapse = "\n")
    asis_output(res)
}
registerS3method("knit_print", "matrix", knit_print.matrix)
```

I am not sure yet if there is a better way to dispatch the printing
function to both matrices and data frames at the same time.

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>

&nbsp;
