---
title: "Selecting columns from data frames in `R`"
date: November 29, 2019
author: Martin Papenberg
---

```{r, include = FALSE}

library(knitr)
knitr::opts_chunk$set(message=FALSE, warning = FALSE) # echo = FALSE
options(scipen = 999, digits = 2)

```

There are quite many alternative ways to select a column from a data 
frame in `R`. In this post I compare some of them with regard to their 
running time. Note that this experiment is of purely scientific 
interest; it does not have any practical consequences because the speed 
of column selection will rarely be the bottleneck of your program.

## The competitors

The `$` selects a single column as a vector; it is the first 
option I was introduced to and I still often use it:

```{r}
sleep$extra
```

The `[[·]]` notation selects a single column from a data frame like, 
just like the `$` notation. However, the column name has to be quoted 
(i.e., we pass a one element vector of type `character`; this has the 
advantage that a function call or a variable can be inserted into the 
double brackets):

```{r}
sleep[["extra"]]
```

The `[·,·]` notation can be used to select multiple columns, but can 
also read just one column as a vector:

```{r}
sleep[, "extra"]
```

We obtain a data frame if we include the argument `drop = FALSE`:

```{r}
sleep[, "extra", drop = FALSE]
```

The `[·]` notation can be used to select multiple columns, but can 
also read just one column as a data frame:

```{r}
sleep["extra"]
```

The function `pull()` from the package `dplyr` can be used to select 
a columns as a vector: 

```{r}
dplyr::pull(sleep, extra)
```

The function `select()` from the package `dplyr` can be used to select 
one or several columns as data frame: 

```{r}
dplyr::select(sleep, extra)
```

## Running time

Using the package `microbenchmark`, I compare the speed of selecting 
single columns. First, I use the `sleep` data frame consisting of 20 
rows; then, I use the `iris` data frame consisting of 150 rows.

```{r}
nrow(sleep)

library(microbenchmark)
microbenchmark(
  sleep$extra, 
  sleep[["extra"]],
  sleep[, "extra"],
  sleep[, "extra", drop = FALSE],
  sleep["extra"],
  dplyr::pull(sleep, extra),
  dplyr::select(sleep, extra)
)
```

I think it is quite interesting that the difference are rather 
substantial. The `$` notation is fastest, about 4 times faster than the 
`[[·]]` notation and more 10 times faster than the `[·]` notation. The 
function `select()` was by far slowest. However, this operation still 
only took about 500 microseconds. As I said, this is definitely not the 
bottleneck of your program. 

Let's repeat the test with a larger data frame:

```{r}
nrow(iris)

microbenchmark(
  iris$Sepal.Length, 
  iris[["Sepal.Length"]],
  iris[, "Sepal.Length"],
  iris[, "Sepal.Length", drop = FALSE],
  iris["Sepal.Length"],
  dplyr::pull(iris, Sepal.Length),
  dplyr::select(iris, Sepal.Length)
)
```

Interestingly, the operations take the same time, even for the larger 
data frame. The rank order in speed remains the same. 

Let's check out a "very large" data frame having 100,000 rows:

```{r}

data <- data.frame(
  col1 = rnorm(100000), 
  col2 = rnorm(100000)
)

nrow(data)

microbenchmark(
  data$col1, 
  data[["col1"]],
  data[, "col1"],
  data[, "col1", drop = FALSE],
  data["col1"],
  dplyr::pull(data, col1),
  dplyr::select(data, col1)
)

```
  
We can state that the speed of reading a column from a data frame is 
not affected by the number of rows!

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>

&nbsp;
