---
title: "Ifelse logic in base `R` and `dplyr`"
date: January 15, 2020
author: Martin Papenberg
---

How to add a column to a data frame using the `ifelse()` function from 
base `R`, creating a categorical vector based on several logical 
conditions:

```{r}
scores <- within(iris, {
  investigate <- ifelse(Sepal.Width > 4, "too-wide", "ok")
  investigate <- ifelse(Sepal.Length < 4.5, "too-short", investigate)
})

table(scores$investigate) # what we got
```

This is how to accomplish the same using the `dplyr` package from the 
popular [tidyverse](https://www.tidyverse.org/):

```{r, message = FALSE, warning = FALSE}
library(dplyr)

scores <- scores %>% 
  mutate(
    investigate = case_when(
      Sepal.Width > 4 ~ "too-wide",
      Sepal.Length < 4.5 ~ "too-short",
      TRUE ~ "ok"
    )
  )

table(scores$investigate)
```

A downside in base `R`: Multiple calls to `ifelse()` are necessary to 
realize more than two categories ("too-wide", "too-short", "ok"). A 
downside in `dplyr`, the default condition (the "else" branch) has to 
use the rather awkward `TRUE ~ "ok"` construction. So: Just use whatever 
pleases you :-)

Inspired by [this blog post](https://thewoodpeckr.wordpress.com/2019/11/22/if-ifelse-had-more-ifs-and-an-else/). 

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>

&nbsp;
