---
title: "Selecting columns from matrices and data frames in `R`"
date: May 17, 2020
author: Martin Papenberg
---

```{r, include = FALSE}

library(knitr)
knitr::opts_chunk$set(message=FALSE, warning = FALSE) # echo = FALSE
options(scipen = 999, digits = 2)

```

<a href="19-columns.html" target="_blank">In a previous post</a>, I 
compared the speed of different data frame column selection operations 
in `R` (the `$` operation performed fastest). In this post, I compare 
the speed of column selection in data frames and matrices.

Using the package `microbenchmark`, I compare the speed of selecting 
single columns. First, I generate some random data and store it as 
matrix and data frame. 

```{r}
N <- 100
mat <- matrix(rnorm(N * 2), ncol = 2)
colnames(mat) <- c("c1", "c2")
df <- data.frame(mat)
```

Now, let's compare how fast columns are selected:

```{r}

library(microbenchmark)
microbenchmark(
  mat[, 1],
  df$c1
)
```

Interestingly, selecting a column from a data frame is a little bit 
faster than from a matrix. However, most interestingly, the discrepancy 
grows with increasing number of elements in the data frame:

```{r}
N <- 10000
mat <- matrix(rnorm(N * 2), ncol = 2)
colnames(mat) <- c("c1", "c2")
df <- data.frame(mat)
microbenchmark(
  mat[, 1],
  df$c1
)
```

```{r}
N <- 100000
mat <- matrix(rnorm(N * 2), ncol = 2)
colnames(mat) <- c("c1", "c2")
df <- data.frame(mat)
microbenchmark(
  mat[, 1],
  df$c1
)
```

Whereas the speed of column selection is hardly affected for data 
frames, column selection gets considerably slower with matrices. I was 
quite surprised by this finding.

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>

&nbsp;
