---
title: "Is the selection of columns via `$` non-standard evaluation?"
date: May 22, 2020
author: Martin Papenberg
---

```{r, include = FALSE}

library(knitr)
knitr::opts_chunk$set(message=FALSE, warning = FALSE) # echo = FALSE
options(scipen = 999, digits = 2)

```

`R` has this magical thing called [non-standard 
evaluation](http://adv-r.had.co.nz/Computing-on-the-language.html). In this 
post, I investigate whether selecting columns in a data frame via `$` 
corresponds to non-standard evaluation. Note that I use the following personal 
working definition of non-standard evaluation:

> An `R` function performs non-standard evaluation if (at least) one of its 
arguments does not by itself evaluate to an `R` object.

This definition is incomplete and does not capture the "technical" essence of 
non-standard evaluation, but I guess it is useful from the viewpoint of an `R` 
user. For example, in the following example, the function `subset()` performs 
non-standard evaluation to select rows in the `sleep` data set, because the code 
`ID == 1`, passed as the second argument, does not evaluate to an `R` object.

```{r}
subset(sleep, ID == 1)
```

`ID` is not itself a variable, but only a column in the data frame `sleep`;
therefore, `ID == 1` most likely throws an error if evaluated by itself in the 
`R` console. However, when passed as an argument to `subset()`, `ID == 1` is 
valid `R` code because the function internally deals with it; `subset()` knows 
that `ID` has to be a column of the data frame `sleep` that is passed as the 
first argument.

Using the above definition, we can investigate whether the `$` notation for 
selecting columns in a data frame corresponds to non-standard evaluation. For 
starters, we note that in `R` every operation is a function call, including 
special syntax for data selection (such as `$`, or `[·]`). Moreover, every 
operation can also be expressed in "standard" function notation using round 
brackets. In the case of the `$` notation, the two following calls are 
equivalent:

```{r}
sleep$ID
`$`(sleep, ID)
```

This already answers the original question: The selection of columns via `$` is 
non-standard evaluation because the second argument -- `ID`, the name of the 
column that is selected -- is passed as a name that by itself does not evaluate 
to an `R` object.

In contrast, column selection by `[[·]]` -- which is equivalent to the `$` 
notation -- performs standard evaluation:

```{r}
sleep[["ID"]]
`[[`(sleep, "ID")
```

This function call performs standard evaluation because `"ID"` is an `R` object 
-- a character vector of length 1. Interestingly enough, <a 
href="19-columns.html" target="_blank">column selection via `$` is faster than 
column selection via `[[·]]`</a>; in terms of speed, non-standard evaluation 
beats its standard counterpart.

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>

&nbsp;
