---
title: "`abtest`: A (not so) new `R` package to compute Bayes factors"
date: March 1, 2021
author: Martin Papenberg
output:
  html_document:
    df_print: kable
---

```{r, include = FALSE}

library(knitr)
knitr::opts_chunk$set(message=FALSE, warning = FALSE)
options(scipen = 999)

set.seed(123)

```

I like Bayes factors and I am always thrilled to find out the package 
[`abtest`](https://github.com/quentingronau/abtest) (also see the corresponding 
[arxiv paper](https://arxiv.org/abs/1905.02068)) was released that enables the 
Bayesian comparison of independent proportions. Similarly, the BayesFactor 
package can also be used for the comparison of proportions, hence in this post I 
want to compare these two approaches informally.

*Note: there has been a release of yet another package for binomial
comparisons.*: **multibridge**

- paper: https://psyarxiv.com/qk4cy/;
- package: https://cran.r-project.org/package=multibridge

**Update**: A paper is available that compares some BFs for the comparison of proportions: https://arxiv.org/abs/2108.04909 (A Puzzle of Proportions: Two Popular Bayesian Tests Can Yield Dramatically Different Conclusions). Also see https://psyarxiv.com/z64th (Bayesian Inference for the A/B Test: Example Applications with R and JASP)

More examples on Bayesian proportion comparisons in this Twitter thread:

https://twitter.com/EJWagenmakers/status/1456994146862907400

## Usage

I am using data from one of my papers where I computed a Bayes factor for the 
comparison of proportions, yielding evidence in favor of the null hypothesis. In 
this publication used the [awesome BayesFactor 
package](https://cran.r-project.org/package=BayesFactor), which at the time was 
the only Bayes factor software implementation I was aware of to achieve this 
goal. Now, with the `abtest` package, however, another approach is possible. 

```{r data}

classifications <- matrix(c(195, 190, 
                            68, 67),
                          byrow = TRUE,
                          ncol = 2)
rownames(classifications) <- c("correct", "incorrect")
colnames(classifications) <- c("MC", "DOMC")

percentages <- round(prop.table(classifications, margin = 2), 4) * 100
```

In my research, I compared two tests with regard to their classification 
accuracy. Both tests performed rather 

```{r}
library(abtest)

data <- list(
  y1 = classifications["correct", "MC"],
  n1 = sum(classifications[, "MC"]),
  y2 = classifications["correct", "DOMC"],
  n2 = sum(classifications[, "DOMC"])
)
bf_abtest <- ab_test(data = data, prior_prob = c(H1 = 0.5, "H+" = 0, "H-" = 0, H0 = 0.5))
print(bf_abtest)

```

In my paper I was interested in the Bayes Factor 0/1, i.e., the degree to which 
the null hypothesis is favored over the alternative, whereas the alternative is 
not specified and either test could outperform the other with regards to the 
classification performance. We can extract this Bayes factor as follows:

```{r}
1 / ab$bf$bf10
```

The Bayes factor we reported in the paper was however the Bayes factor computed 
by the BayesFactor package, which is the Gunel & Dickey Bayes factor (Gunel & 
Dickey, 1974). Using the default setting, it is computed as follows:

```{r}

library(BayesFactor)
bf_bayesfactor <- contingencyTableBF(
  classifications, 
  sampleType = "indepMulti", 
  fixedMargin = "cols"
)
bf_bayesfactor

```

The Bayes factor itself is extracted as follows:

```{r}
1 / extractBF(bf_bayesfactor)$bf
```

So this Bayes factor favors the null hypothesis to a lesser degree as compared 
to the Gunel & Dickey Bayes factor. Still, both Bayes factors qualitatively 
agree with each other. 

## Simulation

In a very small simulation, I compared these two Bayes factors with each other.

```{r}



```


## References 

Gronau, Q. F., & Wagenmakers, E. J. (2019). Informed Bayesian inference for the 
A/B test. arXiv preprint arXiv:1905.02068.

Gunel, E., & Dickey, J. (1974). Bayes factors for independence in contingency 
tables. *Biometrika*, 61, 545–557.

Kass, R. E., & Vaidyanathan, S. K. (1992). Approximate Bayes factors and 
orthogonal parameters, with application to testing equality of two binomial 
proportions. *Journal of the Royal Statistical Society: Series B 
(Methodological)*, 54(1), 129-144.

Papenberg, M., Diedenhofen, B., & Musch, J. (2019). An experimental validation 
of sequential multiple-choice tests. *Journal of Experimental Education*. 
Advance Online Publication. https://doi.org/10.1080/00220973.2019.1671299.
Preprint available from https://osf.io/ycaex/.

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>

&nbsp;
