---
title: "What Bayes factor should you expect?"
date: December 18, 2017
author: Martin Papenberg


bibliography: "lit.bib"
csl: "apa.csl"

---

The usage of Bayesian statistics, and in particular the usage of Bayes
factors is on the rise in Psychology. Bayes factors allow to quantify
the relative evidence that data provide for two different
hypotheses. They therefore lend evidence to one hypothesis over another,
which is different from traditional significance testing relying on
*p*-values; here, "too extreme" data are taken as a decision ground to
discard a null hypothesis (the null hypothesis usually assumes there is
no effect of interest, e.g. that two group means do not differ). 

Bayes factors have been advocated in Psychology as an alternative to
statistical hypothesis testing [@wagenmakers2017; xx]. In particular,
several "default" Bayes factors have been proposed that compare an
alternative hypothesis assuming an effect to a null hypothesis assuming
no effect [e.g. ...]. Such Bayes factors answer the question as to how
much a specified alternative hypothesis describe the data better than
the null hypothesis. In general however, it is not necessary that one of
the competing hypotheses is a null hypothesis. When obtaining a Bayes
factor of size 5, they are 5 times more likely under the alternative
hypothesis than under the null hypothesis; you may say that the data
favor the alternative hypothesis by a factor of 5. You may not say that
the alternative hypothesis is 5 times more likely than the null
hypothesis. I will not give a detailed introduction to Bayes factors
here, but there is a variety treatments you can consult if you are
interested [e.g., ...]

The general appraisal of Bayes factors seems to reach from "*they will
save statistical inference in Psychology from evil p-values*" to "*it is
next dangerous statistic that will be applied mindlessly and lead to
bogus interpretations of data*". However one may think of Bayes factors,
it is probably useful to know how they behave under different settings
-- given their increasing occurrence in Psychological research. In this
post I therefore describe what Bayes factors you might expect when
varying the sample size and the effect size. As is done often, I will
stick to the example of the independent sample *t*-test that compares
two means, and I will use the "default" Bayes factor proposed by
@rouder2009, which is implemented in the R package BayesFactor
[@R-BayesFactor]. This is also the Bayes factor used in the statistics
program JASP when doing the Bayesian *t*-test. Note that my evaluation
of Bayes factors is limited to this specific model.

## The default Bayes factor

## The distribution of Bayes factors

The following plot illustrates the distribution (what that means in the
current case will be explained) of the default Bayes factor for sample
sizes 50 - 1000 and effect sizes d = 0 (i.e. the null hypothesis is
true), d = 0.2 (a small effect), d = 0.5 (a medium effect) and d = 0.8
(a large effect). Null and small effect share the same scaling of the
y-axis; medium and large effect also share the same scaling but it is
different to the scaling of the y-axis of null and small effect. The
data were generated in a simulation assuming normally distributed
scores.

So, how do I understand the plots?


```{r,  message = FALSE, warning = FALSE, echo = FALSE, results="hide", fig.width = 8}
# set knitr options
knitr::opts_chunk$set(message=FALSE, echo=FALSE, warning=FALSE)

source("./rsource/functions.R")

get(load("./data/defBF.data")) # check out names(defBF)

quants <- c(0.2, 0.5, 0.8)
quantiles_02 <- BF_quantiles(defBF$BF02, quants)
quantiles_0  <- BF_quantiles(defBF$BF0,  quants)

## plot null effect and small effect d = 0.2
par(mfrow=c(1,2))
plot_BF_quantiles(quantiles_0, c(1/10, 1/3, 1, 3, 10),
                  ylim=c(-3, 4), main="d = 0", axis="log")
plot_BF_quantiles(quantiles_02, c(1/10, 1/3, 1, 3, 10),
                  ylim=c(-3, 4), main="d = 0.2", axis="standard")
```

```{r,  message = FALSE, warning = FALSE, echo = FALSE, results="hide", fig.width = 8}

quantiles_08 <- BF_quantiles(defBF$BF08, quants)
quantiles_05 <- BF_quantiles(defBF$BF05, quants)

## plot null effect and small effect d = 0.2
par(mfrow=c(1,2))
plot_BF_quantiles(quantiles_05, c(1/10, 1/3, 1, 3, 10),
                  ylim=c(-3, 6), main="d = 0.5", axis="log")
plot_BF_quantiles(quantiles_08, c(1/10, 1/3, 1, 3, 10),
                  ylim=c(-3, 6), main="d = 0.8", axis="standard")
```

There are other treatments relevant to this question:

Sequential Hypothesis Testing With Bayes Factors: Efficiently Testing
Mean Differences (Schönbrodt, 2015) 

Dance of the Bayes factors (Lakens, 2016)
http://daniellakens.blogspot.de/2016/07/dance-of-bayes-factors.html

# References
