---
title: "the anticlust blog"
date: May 03, 2024
author: Martin Papenberg
---

## Using the Mac builder

```{r, echo = FALSE}
set.seed(6789)
options(scipen = 999)
```

The mac builder (https://mac.r-project.org/macbuilder/submit.html) can be used as a testing ground before submitting packages to CRAN. It has "the same setup" as the M1Mac test station on CRAN. My package `anticlust` has produced a segfault on M1Mac on CRAN and therefore I was keen to be able to reproduce the issue on a test station for the purpose of debugging. Unfortunately, it seems that the Mac builder does not have "the same setup" (see [here](https://www.stats.ox.ac.uk/pub/bdr/M1mac/README.txt) maybe) as the M1Mac test station on CRAN, given that the segfault could not be reproduced. 

Using the Mac builder with the release version of R is quite straight forward. You can just submit your package as the tar ball that is generated via `R CMD build`. However, the M1mac on CRAN runs a development version of R ("R unstable") and we have to select the development version to obtain a more realistic test of the CRAN settings. The bad news is that is is very likely impossible for the vast majority of R packages. 

The dependencies of the to-be-tested package need to be submitted manually by the user because there are no packages for R unstable on CRAN that can be accessed instead (unlike for the [win builder](https://win-builder.r-project.org/), it seems). Because the package dependencies need to be installed and tested on the Mac builder themselves, we are also required to upload their dependencies, and so on---even all the "suggested" dependencies recursively. This leaves us with more than 1500 dependencies of `testthat` alone, which is used by many packages for unit tests:

```{r}

imports <- tools::package_dependencies("testthat", recursive = TRUE)
suggests <- tools::package_dependencies("testthat", recursive = TRUE, which = "Suggests")

length(unlist(imports)) + length(unlist(suggests))

```

Uploading all of these packages exceeds the allowed data limit and is therefore impossible.

`anticlust` has few dependencies, I thought, but some of the suggested dependencies are heavy. `anticlust` also used `testthat`, so I switched to [`tinytest`](https://github.com/markvanderloo/tinytest), which has 0 dependencies, so I was able to please the Mac builder (I also think in general this was a good switch and I keep `tinytest` from now on). For vignette building, `anticlust` also has suggested dependencies on `knitr` and `rmarkdown`, which themselves have a lot of additional dependencies. So I generated a version of anticlust without vignettes for testing, which is quite straight forward -- I just removed `knitr` and `rmarkdown` (and also `palmergenguins`, which is only used in a vignette) from the DESCRIPTION and removed the "vignettes" directory. In total, this only left `anticlust` plus an additional 5 packages that had to be uploaded. Well, at least after I generated a version of `RANN` that excludes the `testthat` dependency (by removing `testthat` from the DESCRIPTION and the "tests" directory).

Unfortunately, the Mac builder still did not reproduce the CRAN issue. But well, it is at least one way to test your package on a Mac before CRAN submission if you---as I---do not have a Mac.

---

Last updated: `r Sys.Date()`

### <a href="index.html">Back to the front page</a>
